---
title: "Data Visualization"
author: "Leili Mortazavi"
date: "8/3/2020"
output: 
  pdf_document:
    toc: TRUE
    toc_depth: 2
---


\newpage

Many people have contributed to developing and revising the tutorial material over the years: 

Anna Khazenzon 
Cayce Hook
Paul Thibodeau
Mike Frank
Benoit Monin
Ewart Thomas
Michael Waskom
Steph Gagnon 
Dan Birman
Natalia Velez,
Kara Weisman
Andrew Lampinen
Joshua Morris
Yochai Shavit
Jackie Schwartz
Russ Poldrack

\newpage


In this modulw, you'll learn the basics of how to plot with `ggplot2`,  
a powerful package that enables the construction of highly   
customizable and visually pleasing graphs. 


# Setting up

First, let's import our dataset.

Even though we already have the dataset loaded in our environment,
we want each script to be standalone.
So let's first clea out orur workspace with   
`rm(list=ls())`. Enter this command into your console.
(Remember Console is typically the bottom left window).
You could also click on the broom icon in you environment pane (top right).

Then, we need to load the packages that we want to use in this script.
We'll need `tidyverse`, which includes `ggplot2`.

```{r load packages}
library(tidyverse)
```

Now import the tidy dataset that we saved out in the last module.

```{r import data}
data <- read_csv("../data/prepost_tidy.csv")
```

```{r inspect data}
head(data)
```


# How does ggplot work?

Like dplyr, ggplot takes an object and adds "layers" to it.   
However, while dplyr use piping ("%>%") ggplot uses the plus sign ("+"). 

## Basic layers

Let's see what that means.

```{r ggplot space}
ggplot()
```

As you see, ggplot alone (i.e., without any arguments)
provides a space for your plot to sit in.

Now let's add axes.

```{r ggplot layer 1}
ggplot(data = data,
       aes(x = condition, y = diff))
```
This just sets up the axes, but doens't actually put any data on the plot.
To do that, let's add a "layer" with `+`

```{r ggplot layer 2}
ggplot(data = data,
       aes(x = condition, y = diff)) +
  geom_point()
```

Let's experiment with some more layers:

```{r ggplot more layers}
ggplot(data = data,
       aes(x = condition, y = diff)) +
  geom_point() +
  geom_boxplot()
```

ggplot is very versatile. It can take many many layers, 
and each layer can take some arguments to customize it.
But this is really the skeleton it what it does.

Basically, the first argument is 'data=' so that ggplot knows what 
dataset to work with.  
Then, 'aes()' tells ggplot what the basic "aesthetics" of the plot are.   
These include the x-axis, y-axis (when relevant) and color/fill settings   
(if we want lines or bars in different colors in their outlines or filled   
with different colors respectively).

Remember piping `%>%`?
Let's pipe `data` into ggplot. 

```{r pipe in data into ggplot}
# take the dataset and feed it into ggplot's data argument
data %>% 
  ggplot(data = ., 
         aes(x = condition, y = diff)) +
  geom_point() +
  geom_boxplot()
```

We may want to save this plot into a variable.

```{r assign plot to a variable}
figure1 <- data %>% 
  ggplot(data = ., 
         aes(x = condition, y = score)) +
  geom_point() +
  geom_boxplot()

# just type the name of the plot to see it in here
figure1
```

# Saving out a plot

This plot is now saved in our workspace as figure1. 
But what if I want to save it on my computer as a .png ?

```{r save plot as a png}
ggsave(filename = "figure1.png",
       plot = figure1,
       path = "../figures/.")

# if your plot's proportions look weird when saved,
# you can customize its dimensions as below
ggsave(filename = "figure1.png",
       plot = figure1,
       path = "../figures/.",
       width = 7, height = 4)
```

## Geom objects 

We added points and a line with `geom_point()` and `geom_boxplot()`
But we could also add bars and boxplots and many other types of plots.
To do this, ggplot uses `geom` objects, such as `geom_smooth()`

```{r boxplot}
data %>% 
  ggplot(data = ., 
         aes(x = age, y = score)) +
  geom_smooth(method = "lm")
```

Or a histogram with:

```{r histogram}
data %>% 
  ggplot(data = ., 
         aes(x = score)) +
  geom_histogram()
```

Or a bar plot:

```{r barplot}
data %>% 
  ggplot(data = ., 
         aes(x = score)) +
  geom_bar()
```


## Using facets and colors

We may want to get a sense of how men and women are different. 
Let's take our initial plot, and show men and women with a different color.

```{r use color in ggplot}
data %>% 
  ggplot(data = ., 
         aes(x = condition, y = score, color = gender)) +
  geom_point() +
  geom_boxplot()
```


Now let's plot pretest and posttest separately.

```{r facets in ggplot}
data %>% 
  ggplot(data = ., 
         aes(x = condition, y = score, color = gender)) +
  geom_point() +
  geom_boxplot() +
  facet_grid(. ~ prepost)

# data %>% 
#   ggplot(data = ., 
#          aes(x = condition, y = score, color = gender, fill = gender)) +
#   geom_point() +
#   geom_boxplot() +
#   facet_grid(. ~ prepost)
```

(What's that `~` about? It means row by column.)


There's many more things you can do with ggplot.

* Here's a [reference for `ggplot()`'s options](https://ggplot2.tidyverse.org/), 
with examples. 

To get a sense of all the possibilities, checkout: [R graoh gallery](https://www.r-graph-gallery.com/)

# A few extra resources 

* Take a moment to [learn how to write R code with good style](http://adv-r.had.co.nz/Style.html)--that is, code that's readable and pleasant to look at.

* Check out [dplyr's documentation](https://cran.rstudio.com/web/packages/dplyr/vignettes/introduction.html) for more info on what you can do with dplyr (and how to do it).

* For a reference guide to [plotting data with ggplot, try the R Cookbook](http://www.cookbook-r.com/Graphs/). Personally, I've found the author's published book, the _R Graphics Cookbook_, particularly helpful.  


* For a general reference guide to [using R for data science](http://r4ds.had.co.nz/index.html), i.e. all the sorts of stuff we've been doing, look at Grolemund & Wickham's eponymous [book](http://r4ds.had.co.nz/index.html). (The whole book is online!) 

* `tidyr` is a handy package for [reshaping data from wide to long format and back](https://blog.rstudio.org/2014/07/22/introducing-tidyr/).

* `broom` is a handy package for [turning default output from regressions and t-tests into organized data frames](https://github.com/dgrtwo/broom).

* It's possible (and can be convenient) to [set up a git pane and make commits directly from R.](http://happygitwithr.com/) 

