---
title: "Data Wrangling"
author: "Leili Mortazavi"
date: "8/3/2020"
output:
  pdf_document:
    toc: TRUE
    toc_depth: 2
---


\newpage

Many people have contributed to developing and revising the tutorial material over the years: 

Anna Khazenzon 
Cayce Hook
Paul Thibodeau
Mike Frank
Benoit Monin
Ewart Thomas
Michael Waskom
Steph Gagnon 
Dan Birman
Natalia Velez,
Kara Weisman
Andrew Lampinen
Joshua Morris
Yochai Shavit
Jackie Schwartz
Russ Poldrack

\newpage
# Note on relative paths vs. absolute paths

Before we dig into working with data in R, let's talk about some good
practices about how to write down your paths.

If you look at the top of your console panel, you can see your path.
(i.e., where you are in the eyes of RStudio).

If you open up an R script in a new session of RStudio, the path will
automatically be set to where your R script is saved. 

If you already have other R scripts open in RStudio and want to work 
on a different script, you may need to change your path to where that 
script is saved.

To do this run:

```{r set working directory, eval = FALSE}
setwd("~/PATH/TO/CURRENT/SCRIPT")
```

Why bother??

If you can keep a tidy directory structure, then you can write a 
**relative path** for importing and exporting data and plots. 
Not only will this make your coding easier, it will make your script 
reproducible (for your future self and for others). 
So if you move your project directory, or if someone else wants to 
work on it on their own device, the script will work smoothly, 
without the need to change all the paths every time.


So now that we've started working on a new script, which is saved 
in a different folder, let's use relative path to go there:

```{r set working directory with relative path}
setwd("../Part2")
```

As you see now, all of us can now use this relative path within our
R_basic folder, regardless of having saved the bootcamp files 
in different locations. 

Just to convince you a little more, if I wanted to use absolute path
this is how it would look like:
`setwd("~/STANFORD/teaching/summerStats_2020/modules/R_basic/Part2")`

Ugly huh?!
Plus, yours would probably be completely different up until the 
"modules" part. 

# Importing data

Now, let's use relative path to import the albums dataset, 
sitting in you data folder.

```{r import .csv}
data <- read.csv("../data/albums.csv")
```


# Inspecting data

```{r check data}
View(data)
# note that View has to be capitalized!

head(data)
```

This type of data is called a dataframe.

Let's formally check its type to see what R thinks of it:

```{r dataframe as a variable type}
class(data)
```

Get a list of all the column names:

```{r check column names}
colnames(data)
```

Remember we said R tries to detect data types? 
Even though the whole dataset is a data frame, each column may
be numeric, character, etc. Let's check:

```{r check type of a specific column}
# use your indexing skills
# select a column based on its index (i.e., first column):
class(data[,1])

# select a column based on its name, using `$` as follows:
class(data$album)
```

Note: The TAB key is your friend! 
Use it to make your life easier, 
AND to decrease the chances of making a typo, 
which will result in an annoying hard-to-detect error.

We can use the summary function of a data frame to get
some summary values on each column:

```{r summary of each column}
summary(data)
```

```{r check for NA values}
# on the entire data frame
is.na(data)

# on a specific column
# TRY FOR YOURSELF!
# Hint: use your indexing skills.



# is.na(data$year)
```

# Manipulating data

When we analyze and visualize the data, we often need to manipulate data in some ways. 
For instance, we may need to select only 2 of the variables, 
or we may need to take the mean and standard deviation of a subset of our observations,
e.g., separately for males and females.

## Tidyverse

To do these, and many more operations, there's a set of packages that come in very handy.
That is, tidyverse (by Hadley Wikham): a collection of packages that make  
transforming data a lot easier!


```{r install and load tidyvers}
# install.packages("tidyverse")

# load it
library(tidyverse)
```

**Note**: when we write scripts, it is conventional to 
load all the required packages on top of the script.

## Selecting and subsetting

```{r select some of the variables}
select(.data = data, year, artist)
```

```{r select and save in a new dataframe}
data_1 <- select(.data = data, year, artist)
head(data_1)
```

```{r subset only some observations}
data_2 <- subset(data_1, year > 1995)
head(data_2)
```

## Piping

It's quite inefficient to create new variables every time we want to perform
some operation. What if we want to perform several steps (which we often do)?

Let's learn a really nice feature of the tidyverse package that will make 
coding in R much easier. 

This is called ** piping**. ` %>% `
Piping will allow us to do several operations in one big chain. 

```{r piping}
# select and subset in one go

data_select <- data %>%        
  # means: take the dataframe and feed it into the next line
  select(year, artist) %>%     
  # means: select these two columns from the "data" dataframe, 
  # and feed that into the next line
  subset(year > 1995)          
  # means: subset some of the rows and save the result in the "data_select" dataframe
head(data_select)
```
## Making new variables

```{r make a new variable}
data %>% 
  mutate(sales = sales_millions * 1000000)
```

We learned 3 important and very common functions of tidyverse: 
`select()`, `subset()`, and `mutate()`

Let's practice now.

```{r practice}

```


## Grouping and summarizing

dplyr (part of tidyvers) can also be used to quickly summarize data by 
different grouping variables. 

In this dataset, who's the bestselling artist?

To find out, we group by `artist`, calculate each artist's total album   
sales with `summarize()`, and save the result in `result`:

```{r more_dplyr}
result <- data %>%                              
  # group the data by `artist`
  group_by(artist) %>%                             
  # add up the `sales` for each artist and save in `total_sales`
  summarize(total_sales = sum(sales))

head(result)
```


```{r practice summarize, eval = FALSE}
# setting eval to false in the brackets tells R not to run this code when knitting
# APPLY YOUR KNOWLEDGE:
# How would you change this code if we instead wanted to compare  
# the total sales, in millions, of solo artists vs. group artists?

result <- data %>%                                
  mutate(sales = sales_millions * 1000000) %>%     
  group_by(artist) %>%                             
  summarize(total_sales = sum(sales)) %>%          
  arrange(desc(total_sales)) 
















# ANSWER:
# result <- albums %>%                                
#  group_by(solo) %>%                             
#  summarize(total_sales = sum(sales_millions))

# you can also group by multiple variables
result <- albums %>%
  group_by(artist, solo) %>%
  summarize(total_sales = sum(sales_millions)) %>%
  arrange(desc(total_sales))
```


# Tidy data

Tidy data is long data, where every row is one observation.
If there are multiple observations per subject, then we have 
multiple rows per subject,
but then have another variable (i.e., column) denoting some 
aspect of those observations.

Let's work through an example.

Import a dataset from your data/ folder.
The file is called "prepost.csv".

```{r import new data}
# practice importing csv files







# ANSWER:
data_wide <- read_csv("../data/prepost.csv") 
```

```{r inspect your data}
# first in matrix view: 

# top 5 rows:

# names of columns:







# ANSWER:
# first in matrix view: 
# View(data_wide)

# top 5 rows:
# head(data_wide, 5)

# names of columns:
# colnames(data_wide)
```

As you can see, this dataset is not in tidy format, 
because we have two observations per row (i.e., pretest and posstes).

Remember that tidy data is long data, where every row is one observation.
If there are multiple observations per subject, then we have multiple 
rows per subject, but then have another variable (i.e., column) denoting 
some aspect of those observations.

Here, we want to make a variable that only denotes whether the score was
from pretest or from posttest. Let's call this variable "prepost".

Then we want to put the score into another column. 
Let's call this variable "score".

So how do we do that?

Here's a function in tidyverse that takes care of this operation very gracefully:
`pivot_longer()`

```{r pivot longer}
data_long <- data_wide %>% 
  pivot_longer(cols = c(pretest, posttest),   
               # means: select the columns that include the scores
               names_to = "prepost",         
               # means: take the name of these two columns and put 
               # them into a new variable called "prepost"
               values_to = "score"           
               # means: take the values of these two columns and put them into a new variable called "score" 
               )
```

Now let's check out what pivot_longer() did. 

```{r check tidy data}
data_long %>% 
  head()

# How many rows did the wide data have?


# How many rows does the long data have? 

```

As we work through the visualization modules and statistical analyses,
you'll see the benefits of having tidy data. 

# Saving out wrangled data

After we manipulate out data in ways that's ready for analyses or visualizations, 
we may want to save it out in a csv file, so that we don't need to wrangle it again.


```{r save out data}
write_csv(x = data_long, 
          path = "../data/prepost_tidy.csv")
```


--------------------

For extra material on the following, see module **BasicR_Part2_dataWranglingExtra.Rmd**

- creating dataframes in R
- manipulating data using base R 