---
title: "Data wrangling (2019)"
author: "stats crew from before 2020"
date: "8/6/2020"
output: pdf_document
---

EXTRAS

# More basics about data frames

Most often when doing statistics with R, you'll be using *data frames.*   
Technically, a data frame is a list of vectors.   
It's similar in appearance to a basic spreadsheet in Excel or SPSS,   
where you might represent your study variables in   
columns and your observations in rows. 

It's possible to make a data frame from scratch with the base R function   
`data.frame()` (though in practice, you will probably not do this often):
We'll also be showing you more effective ways to code with the  
tidyverse (by Hadley Wikham): a collection of packages that make  
transforming data a lot easier!

```{r dataframe_intro}

albums = data.frame(
  album = c("Speakerboxx/The Love Below",
            "Life After Death",
            "All Eyez on Me",
            "Licensed to Ill",
            "Stankonia",
            "Ready to Die",
            "R U Still Down? Remember Me",
            "Ill Communication"),
  artist = c("Outkast",
             "The Notorious B.I.G.",
             "2Pac",
             "Beastie Boys",
             "Outkast",
             "The Notorious B.I.G.",
             "2Pac",
             "Beastie Boys"),
  year = c(2003,
           1997,
           1996,
           1986,
           2000,
           1994,
           1997,
           1994),
  sales_millions = c(11.4,
                     10.2,
                     9,
                     9,
                     4,
                     4,
                     4,
                     3),
  solo = c(FALSE,
           TRUE,
           TRUE,
           FALSE,
           FALSE,
           TRUE,
           TRUE,
           FALSE)
  )
albums



#### OR ####

album <- c("Speakerboxx/The Love Below",
            "Life After Death",
            "All Eyez on Me", 
            "Licensed to Ill", 
            "Stankonia",
            "Ready to Die",
            "R U Still Down? Remember Me",
            "Ill Communication")
artist <- c("Outkast",
             "The Notorious B.I.G.",
             "2Pac", 
             "Beastie Boys",
             "Outkast", 
             "The Notorious B.I.G.",
             "2Pac", 
             "Beastie Boys")
year <- c(2003,
           1997,
           1996,
           1986,
           2000,
           1994,
           1997,
           1994)

sales_millions <- c(11.4,
                     10.2,
                     9,
                     9,
                     4,
                     4,
                     4,
                     3)
solo <- c(FALSE, 
           TRUE,
           TRUE,
           FALSE,
           FALSE, 
           TRUE,
           TRUE,
           FALSE)

albums <- data.frame(
  album,
  artist,
  year,
  sales_millions,
  solo
) # can also use tibble() instead of data.frame from tidyverse
albums
```


### Subsetting data

Every time you want to do something with a column in a data frame,  
you use the `$` to specify the data frame and column you want to act upon.  
Since these columns are really vectors,  
what we've learned about brackets and functions also applies here.   
So, we can use brackets to return all the observations where   
a certain condition is true: 

```{r get_a_column}
albums$artist # pull out a specific vector

albums$artist[albums$sales_millions > 10] # subset a variable based
# on a logical statement. In this case, returns the artisits
# who issued albums with over 10 million in sales.

#### OR ####
# install.packages("tidyverse")
library(tidyverse)
# let's say you want to creat a new dataframe with artists who issued 
# albums with over $10 mil in sales
albums_overten <- albums %>%
  filter(sales_millions > 10)

```

You can also use brackets directly on the dataframe,  
but since it has two dimensions, the brackets contain a comma.  
Whether you put numbers before or after the comma specifies whether  
you want to do stuff to your data frame's  *rows* or *columns*: 
```{r}
albums[1, ] # returns the first row of `albums`
albums[, 1] # returns the first column 

albums[1:3, 2:4] # returns a new dataframe containing rows 1-3 and columns 2-4 
# generally however, I'd recommend to identify the column or row by name rather than the specific placement in the df
```

You can use this to get rid of variables:
```{r}
albums[, -3] 
# returns a new dataframe with the original third column of `albums` removed
albums[-3, ] 
# returns a new dataframe with the original third row of `albums` removed
```

### Creating new variables

You can add new columns (i.e. variables) to the data frame with `$`: 

```{r expand_data frame}
albums$sales <- albums$sales_millions*1000000
albums$language <- "English"
# if you enter a single value rather than a vector,
# every observation will be set to that value

albums

#### OR ####

albums_new <-
  albums %>%
  mutate(
    sales = sales_millions*1000000,
    language = "English"
  )


# APPLY YOUR KNOWLEDGE:
# Write a line of code to add a new column to `albums`
# called `years_since_release` and populate it with the number of years 
# it's been since the album came out. 






















# You should have written something like: 
# albums$years_since_release <- 2019 - albums$year
# or 
# albums_new <-
#   albums_new %>%
#   mutate(years_since_release =
#            2019 - year)

```

### Recoding variables

Say we're interested in making a new grouping variable to represent  
the decade an album came out ("80s", "90s", or "00s").   
It's possible to recode variables using brackets:

```{r}
# First make an empty placeholder for the new variable:
albums$decade <- NA

# Then set each of the values according to whatever subsetting criteria are relevant:
albums$decade[(albums$year >= 1980) & (albums$year < 1990)] <- "80s"
albums$decade[(albums$year >= 1990) & (albums$year < 2000)] <- "90s"
albums$decade[(albums$year >= 2000) & (albums$year < 2010)] <- "00s"

#### OR ####
## let's just glimpse first to see if year is the data class we expect
glimpse(albums_new)
albums_decade <-
  albums_new %>%
  mutate(decade =
           ifelse(
             year >= 1980 & year < 1990,
             "80s",
             "NA"
             ),
         decade =
           ifelse(
             year >=1990 & year < 2000,
             "90s",
             decade
           ),
         decade =
           ifelse(
             year >=2000 & year < 2010,
             "00s",
             decade
           )
             )
### OR ###
##case_when is also very useful to avoid having so many ifelse arguments
#this is from the tidyverse library  
```

This can be confusing to look at at first, so I'll give another example. 

Say I want to add a variable `feud` whose value depends on whether  
the artist was involved in the [90s East Coast-West Coast hip hop rivalry](https://en.wikipedia.org/wiki/East_Coast%E2%80%93West_Coast_hip_hop_rivalry).   
That is, I want `feud` to be "TRUE" if the artist is   
The Notorious B.I.G. or Tupac, and "FALSE" otherwise. 

Here's one way to do it:

```{r}
albums$feud <- FALSE
albums$feud[albums$artist == "2Pac"] <- TRUE
albums$feud[albums$artist == "The Notorious B.I.G."] <- TRUE


#### OR ### 

albums_feud <-
  albums_decade %>%
  mutate(feud =
           ifelse(artist == "2Pac" | artist == "The Notorious B.I.G.",
         TRUE,
         FALSE)
  )


```

```{r}
# APPLY YOUR KNOWLEDGE:

# Try making a new variable called 'born_yet' which 
# is set to TRUE if the album came out after you were born, and FALSE otherwise.












# My answer:
# albums$born_yet <- FALSE
# albums$born_yet[albums$year >= 1989] <- TRUE
# or
# albums_bornyet <-
#   albums_feud %>%
#   mutate(
#     born_yet =
#       year >= 1991,
#     TRUE,
#     FALSE
#   )

```


### A note on `$`

Remember, you have to use `$` when referring to a variable in a data frame. 

Try `mean(sales_millions)` or even `albums$artist[sales_millions > 10]`.   
Why don't these work?


Look back over at your Environment pane. There's now a category,   
Data, that contains `albums`. 
There's also the Values category, which contains all of the other   
random objects we've made throughout this lesson. 

The object `sales_millions` only exists in the data frame `albums`,   
rather than in your broader 'Global Environment'.

It's useful to keep variables inside their own data frames,  
because it enables you to, e.g., simultaneously work on multiple  
datasets that contain variables with the same name   
(`albums$sales` vs. `pizza$sales`).

-------------------------


# Where do the numbers go? (Part II)

Right now, our data frame only exists in R's memory,  
and its contents depend exactly upon which lines of code  
we've run during this R session (and in what order). 


What if we wanted to save `albums` to share with a friend  
who only has Excel or SPSS? 

R can write data frames to .csv files, which are easily opened by   
many different programs and are a useful format in which to share data. 

To begin, we first have to tell R where to save files.

### Setting your working directory
Whenever you use R, you're in some *working directory* (essentially a folder),  
where R will default to saving files   
(and searching for them when you try to open a file). 

Type `getwd()` to discover your current working directory.


To set your working directory, type `setwd()` with the   
file path in quotation marks inside the parentheses, e.g.

```{r setwd_desktop}
#setwd("~/R_workshop_materials")
```
(You can set your working directory by hand through   
*Session > Set Working Directory > Choose Directory*.) 

(In the future, you should set your working directory to a   
project-specific folder with `setwd()` whenever you start   
working on a new project in R. If you leave it as "~/Desktop" things  
will get cluttered quickly.) 

### Writing a .csv

Now we know where our file will be saved, let's `write.csv()`.  
The tidy way is `write_csv()` which I will be using instead.
The first argument is the R data frame you want to save,   
the second argument is the filename you want to save it under,   
and the third argument specifies that we don't want to save row names. 

```{r}
# you can also set your path to a variable name, e.g.
write_csv(albums_bornyet,
          "albums.csv", 
          append = FALSE,
          col_names = TRUE)
```

Go find `albums.csv` and open it up in Excel.   
Make some changes to it, save it, and close the file.

Now click on `albums` in the Environment pane.   
Do your changes to the document show up in R?

```{r }
#









#
```

Now run:
```{r where_do_the_numbers_go}
albums$artist <- c(
  "Rihanna",
  "Beyonce",
  "Taylor", 
  "Britney", 
  "Adele", 
  "Cher",
  "Whitney", 
  "Celine"
  ) 
head(albums) # see that we've changed the values for 'artist'

#### OR #####
albums <-
  albums_bornyet %>%
  mutate(
    artist = c(
      "Rihanna",
      "Beyonce",
      "Taylor", 
      "Britney", 
      "Adele", 
      "Cher",
      "Whitney",
      "Celine"
      )
    )
  
```

Reopen the .csv file in Excel. Did this change show up?

```{r }
#









#
```


What R has in its memory is distinct from the original .csv file.   
This is a good thing, since it allows you to manipulate your dataset   
(which inevitably involves making mistakes and/or deleting variables   
or observations) while preserving the original, complete raw data. 

Now replace our data frame, `albums`,   
with the contents of `albums.csv` using `read.csv`:

### Reading from a .csv

```{r readcsv}
albums <- read_csv("albums.csv")
head(albums)
```

The correct artist names have been restored! 


---------------


# Manipulating data frames with dplyr 
Tidyverse includes dplyr, so don't need to install dplyr separately


We just reviewed a bunch of different actions you can take with data frames:

* filtering observations based on some criterion 
* transforming variables
* summarizing data with functions like `mean()`

We can do all of these things (and more) using functions  
from tidyverse including the R package `dplyr`, and we can chain it all together   
with the pipe operator, `%>%`. Pipes free us from having to   
precede every variable with `dataframe$`--instead, each line operates on   
the data frame that was fed into it from the previous pipe.  
Hopefully this will be clearer with examples:

```{r}
# First let's reset the value of `albums`:
albums = data.frame(
  album = c(
    "Speakerboxx/The Love Below",
    "Life After Death",
    "All Eyez on Me", 
    "Licensed to Ill", 
    "Stankonia", 
    "Ready to Die",
    "R U Still Down? Remember Me", 
    "Ill Communication"
    ), 
  artist = c(
    "Outkast", 
    "The Notorious B.I.G.", 
    "2Pac", 
    "Beastie Boys", 
    "Outkast", 
    "The Notorious B.I.G.",
    "2Pac", 
    "Beastie Boys"
    ), 
  year = c(
    2003, 
    1997, 
    1996, 
    1986, 
    2000, 
    1994, 
    1997,
    1994), 
  sales_millions = c(
    11.4,
    10.2,
    9,
    9,
    4,
    4, 
    4, 
    3
    ),
  solo = c(
    FALSE, 
    TRUE, 
    TRUE,
    FALSE, 
    FALSE, 
    TRUE,
    TRUE, 
    FALSE)
  )

# Here we'll do two of the things we did above--transforming variables  
# and filtering out observations--using pipes and dplyr functions this time:

albums2 <- albums %>%                             
  mutate(sales = sales_millions * 1000000) %>%    
  filter(sales > 10000000)                        

# The first line specifies that we're modifying `albums` 
# and putting the result in `albums2`.
# The second line takes `albums` and creates a new variable, `sales`,
# and passes this updated dataframe to the next line.
# The third line takes the dataframe from the previous step, and throws out
# the rows where `sales` is less than or equal to 10,000,000.
# After the final line has finished running, the final dataframe is put in `albums2`.
# No changes are made to the contents of `albums`. 

```

